---
title: "Analyzing Pulse-Chase Designs"
output: 
  html_document:
    toc: true
    df_print: kable
    fig_width: 7
    fig_height: 6
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,comment=NA,R.options=list(width = 95))
```

This vignette will show the most suitable commands to work with pulse-chase data.

Throughout this vignette, we will be using the GRAND-SLAM processed SLAM-seq data set from Herzog et al. 2017 [[1]](https://www.nature.com/articles/nmeth.4435). The data set includes SLAM-seq pulse-chase samples of mouse embryonic stem cells (mESCs) with pulse durations of 0, 0.5, 1, 3, 6, 12, and 24 hours.

```{r}
suppressPackageStartupMessages({
  library(grandR)
  library(ggplot2)
  library(patchwork)
})
```

First we load the data and do the standard preprocessing steps. For more on these initial commands see the [Loading data and working with grandR objects](loading-data.html) vignette.

In this step, we need to modify the sample names in the data, as the names were not processed using systematic naming conventions. If systematic sample names were used, the `rename.sample` parameter would not be necessary. This parameter allows us to specify a function for renaming the samples.
The specified function removes all occurrences of ".chase" and replaces 0.5 with 0_5. For example, "mESC.0.5h.chase.B" becomes "mESC.0_5h.B" and "mESC.0h.chase.C" becomes "mESC.0h.C". The output of the function is the modified vector after these replacements. This enables us to load the data using the design vector.

```{r}
d=ReadGRAND("https://zenodo.org/record/7612564/files/chase_notrescued.tsv.gz?download=1",
            design=c(NA,"Time",Design$Replicate),
            rename.sample=function(v) 
              gsub(".chase","",gsub("0.5h","0_5h",v))
            )
d=FilterGenes(d)
d=Normalize(d)
```

The `FitKinetics` function can be utilized to analyze pulse-chase experiment data and estimate the synthesis and degradation rates of genes. To perform this analysis, the type parameter of the function should be set to "chase". The function requires a grandR object as input, which holds the data from the pulse-chase experiment.

In addition to the required input, the `FitKinetics` function also has several optional parameters that allow for customization of the analysis. For further information on these parameters, please refer to the [Kinetic modeling](kinetic-modeling.html) vignette.

The `FitKinetics` function uses the information in the grandR object and the specified parameters to fit a non-linear least squares model that describe the pulse-chase data. After fitting the models, the function returns a grandR object with an added analysis table that contains the inferred synthesis and degradation rates. 

```{r,warning=FALSE}
SetParallel()
d=FitKinetics(d,time="Time",type="chase")
```

The function `GetAnalysisTable` can be used to retrieve the results from this analysis.

```{r}
head(GetAnalysisTable(d))
```

We can also visualize single genes graphically (this is the same gene as used in Fig 4a in Herzog et al. 2017 [[1]](https://www.nature.com/articles/nmeth.4435)):

```{r}
PlotGeneProgressiveTimecourse(d,gene=Genes(d,"Dnmt3b",regex=TRUE),time="Time",type="chase")
```

The curves represent the fitted model for this gene. Note that we do not specify the gene name directly, since here the gene names were labeled differently, and the `Genes()` function can be used to find the matching label:

```{r}
Genes(d,"Dnmt3b",regex=TRUE)
```


In this step, we compare the half-life values obtained from the `FitKinetics` function with those from Herzog et al. The latter was estimated by observing the decrease in mismatches during the chase period, while our method also considers the GRAND-SLAM-estimated NTR and posterior. To compare, we add Herzog et al's half-life values to our analysis table and create a scatter plot with half-life values on both the x and y axes. The plot clearly shows a correlation between the two sets of results, indicating that the `FitKinetics` results are consistent with those from Herzog et al.

```{r,warning=FALSE}
t=read.delim("https://zenodo.org/record/7612564/files/halflifes.tsv?download=1")
t$Gene=paste0(t$Name,".",t$End)
d = AddAnalysis(data = d,"kinetics",table = t,by="Gene")
PlotScatter(d,`kinetics.Half.life..h.`,`kinetics.Half-life`,
            correlation = FormatCorrelation())+geom_abline()
```

